On-the-fly Algorithms for GKAT Equivalences \thanks{Identify applicable funding agency here. If none, delete this.}

### 2025-01-16 main.tex
minor tweak

In this paper, we will make heavy use of coalgebraic theory, thus it is empirical for us to recall some notions and useful theorems in universal coalgebra.
Given a functor \(F\) on the category of set and functions, a \emph{coalgebra over \(F\)} or \emph{\(F\)-coalgebra} consists of a set \(S\) and a function \(σ_S: S → F(S)\).
We typically call elements in \(S\) the \emph{states} of the coalgebra, and \(σ_S(s)\) the \emph{dynamic} of state \(s\).
We will sometimes use the states \(S\) to denote the coalgebra, when no ambiguity can arise.

@@

\begin{proof}
    We will need to show that \(h(⟨s⟩_{S})\) is the smallest sub-GKAT coalgebra of \(S\) that contain \(h(s)\). 
    By definition of image, \(h(⟨s⟩_{S})\) indeed contain \(h(s)\). 
    By \Cref{thm:hom-(pre)img-preserve-sub-coalg}, \(h(⟨s⟩_S) ⊑ U\), i.e. \(h(⟨s⟩_S)\) is a sub-coalgebra of\(U\).
    Finally, take any sub-coalgebra \(U' ⊑ U\) s.t. \(h(s) ∈ U'\): 
    \begin{align*}
        h(s) ∈ U' 
        & ⟹ s ∈ h^{-1}(U') \\  
        & ⟹ ⟨s⟩_S ⊑ h^{-1}(U') & \text{definition of \(⟨s⟩_S\)}\\  
        & ⟹ h(⟨s⟩_S) ⊑ U' & \text{\Cref{thm:hom-(pre)img-preserve-sub-coalg}}
    \end{align*}
    Hence \(h(⟨s⟩_S)\) is the smallest sub-GKAT coalgebra of \(U\) that contains \(h(s)\).
\end{proof}

@@

Traditionally, live and dead states are defined by whether they can reach an accepting state~\cite{smolka_GuardedKleeneAlgebra_2020}. 
However, we can use induction on the length of the path to show that principle sub-coalgebra \(⟨s⟩_S\) contains exactly the reachable states of \(s\) in any GKAT coalgebra \(S\). 
Thus, the classical definition is equivalent to the following:
\begin{definition}[liveness of states]\label{def:liveness-of-states}
    A state \(s\) is \emph{accepting} if there exists an atom \(α ∈ \At\) s.t. \(δ(s, α) = \accept\); a state \(s'\) is \emph{live} if there exists an accepting state \(s' ∈ ⟨s⟩\); and a state \(s'\) is \emph{dead} if there is no accepting state in \(⟨s⟩\).
\end{definition}
This alternative liveness definition can help us prove important theorems regarding reachability and liveness without explicitly performing induction on traces. 
We show the following theorems as examples:
\begin{lemma}\label{thm:dead-iff-all-reachable-dead}
    A state \(s\) is dead if and only if all elements in \(⟨s⟩\) is dead.
\end{lemma}
\begin{proof}
    \(⟸\) direction is true, because \(s ∈ ⟨s⟩\): if all \(⟨s⟩\) is dead, then \(s\) is dead. 
    \(⟹\) direction can be proven as follows.
    Take any \(s' ∈ ⟨s⟩\), then \(⟨s'⟩ ⊑ ⟨s⟩\) because \(s'\) is the minimal sub-coalgebra that contains \(s'\). 
    Since there is no accepting state in \(⟨s⟩\), thus there cannot be any accepting state in \(⟨s'⟩\), hence \(s'\) is also dead.
\end{proof}

>>>>

In this paper, we will make heavy use of coalgebraic theory, thus we will recall useful notions and theorems in universal coalgebra.
Given a functor \(F\) on the category of set and functions, a \emph{coalgebra over \(F\)} or \emph{\(F\)-coalgebra} consists of a set \(S\) and a function \(σ_S: S → F(S)\).
We typically call elements in \(S\) the \emph{states} of the coalgebra, and \(σ_S(s)\) the \emph{dynamic} of state \(s\).
We will sometimes use the states \(S\) to denote the coalgebra, when no ambiguity can arise.

@@

\begin{proof}
    We will need to show that \(h(⟨s⟩_{S})\) is the smallest sub-GKAT coalgebra of \(S\) that contain \(h(s)\). 
    By definition of image, \(h(⟨s⟩_{S})\) indeed contain \(h(s)\). 
    By \Cref{thm:hom-(pre)img-preserve-sub-coalg}, \(h(⟨s⟩_S) ⊑ U\), i.e. \(h(⟨s⟩_S)\) is a sub-coalgebra of \(U\).
    Finally, take any sub-coalgebra \(U' ⊑ U\) s.t. \(h(s) ∈ U'\): 
    \begin{align*}
        h(s) ∈ U' 
        & ⟹ s ∈ h^{-1}(U') \\  
        & ⟹ ⟨s⟩_S ⊑ h^{-1}(U') & \text{definition of \(⟨s⟩_S\)}\\  
        & ⟹ h(⟨s⟩_S) ⊑ U' & \text{\Cref{thm:hom-(pre)img-preserve-sub-coalg}}
    \end{align*}
    Hence \(h(⟨s⟩_S)\) is the smallest sub-GKAT coalgebra of \(U\) that contains \(h(s)\).
\end{proof}

@@

Traditionally, live and dead states are defined by whether they can reach an accepting state~\cite{smolka_GuardedKleeneAlgebra_2020}.
\begin{definition}[Liveness]\label{def:liveness-of-states}
    A state \(s\) is \emph{accepting} if there exists an atom \(α ∈ \At\) s.t. \(δ(s, α) = \accept\).
    A state \(s₀\) is \emph{live} if there exists an accepting state \(sₙ\) s.t. there is a liveness chain:
    \[s₀ \transvia{α₁ ∣ p₁} s₁ \transvia{α₂ ∣ p₂} s₂ ⋯ \transvia{αₙ ∣ pₙ} sₙ.\]
    A state is \emph{dead} if it is not dead.
\end{definition}
However, besides directly working with this explicit definition, we can also give a more coalgebraic characterization of these properties.
Specifically, by induction on the length of the path from one state to the next, we can show that the principle sub-coalgebra \(⟨s⟩_S\) exactly contains all the reachable state of \(s\) in the GKAT coalgebra \(S\).
Thus we obtain the following coalgebraic characterization of liveness:
\begin{theorem}\label{thm:liveness-principle-sub-coalg}
    A state \(s\) is \emph{live} if there exists an accepting state \(s' ∈ ⟨s⟩\); and a state \(s\) is \emph{dead} if there is no accepting state in \(⟨s⟩\).
\end{theorem}
This alternative liveness definition can help us prove important theorems regarding reachability and liveness without explicitly performing induction on the liveness chain. 
We show the following theorems as examples:
\begin{lemma}\label{thm:dead-iff-all-reachable-dead}
    A state \(s\) is dead if and only if all elements in \(⟨s⟩\) is dead.
\end{lemma}
\begin{proof}
    \(⟸\) direction is true, because \(s ∈ ⟨s⟩\): if all \(⟨s⟩\) is dead, then \(s\) is dead. 
    \(⟹\) direction can be proven as follows.
    Take any \(s' ∈ ⟨s⟩\), then \(⟨s'⟩ ⊑ ⟨s⟩\) because \(s'\) is the minimal sub-coalgebra that contains \(s'\). 
    Since there is no accepting state in \(⟨s⟩\), thus there cannot be any accepting state in \(⟨s'⟩\), hence \(s'\) is also dead.
\end{proof}
